Staged Abstract Interpreters

### 2018-11-16 paper/22_bg_abs.tex
paper

\todo{conditional, use different store?}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

>>>>

\todo{conditional, use different store?}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$ $\sqcup$ $\sigma$_*; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

### 2018-11-16 paper/50_cases.tex
paper

The target language we presented in Section~\ref{bg_lang} is essentially a higher-order functional language.
One fundamental analysis task for functional programs is the control-flow analysis, i,e., determining
which functions will possibly be applied at each call-site. 
The abstract interpreter we used in Section~\ref{unstaged_abs} and Section~\ref{sai} is already 
a store-widened 0-CFA-like abstraction; in last section, we also reviewed AC with 0-CFA.
In this section, based on the existing staged abstract interpreter, we further
develop the staging techniques with control-flow analyses, including recoverying a more precise store model;
different polyvariant analsis, and using staging as a implementation stategy for mixed sensitivity.

@@

\subsubsection{Context/Path/Flow-Sensitivity}

\subsubsection{Mixed Sensitivity}

using different $k$ for $k$-CFA

@@

Beyond control-flow analysis, staging data structures used in the abstract interpreter, such as abstract domain.
It has been shown that specializing abstract domains with respect to the structure of analyzed program significantly 
improves the performance: a recent example is online decomposition of polyhedra \cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143}.
In this section, we present a similar idea for interval abstract domain using an \textit{offline} fashion by staging.

@@

Assignments, Loops, Exceptions

\subsubsection{Specializing Interval Domains}

Intervals, Numerical Analysis

Options: Data-flow analysis, taint analysis

>>>>

The target language we presented in Section~\ref{bg_lang} is essentially a higher-order functional language.
One fundamental analysis task for functional programs is the control-flow analysis, i,e., determining
which functions will possibly be applied at each call-site. 
The abstract interpreter we used in Section~\ref{unstaged_abs} and Section~\ref{sai} is a store-widened 
0-CFA-like abstraction, moreover it is also a pushdown control-flow analysis; 
in last section, we also reviewed AC with finite-state 0-CFA.
In this section, based on the existing staged abstract interpreter, we further
develop the staging techniques with control-flow analyses, including recoverying a more precise store model;
context-sensitive analsis, and using staging as a implementation stategy for mixed sensitivity.

@@

The store-widened analysis we implemented in Section~\ref{sai} uses a single store to approximate
the runtime store. It can be efficently computed in polynomial time together with 0-CFA-like abstraction, 
but sometimes we may desire a 
more precise result that distinguish the final values and stores for different closure targets. 
To achieve this goal,
we need to tweak our abstract interpreter and type instantiation. The answer type @Ans@ is changed to
a set of @VS@s where a @VS@ is a pair of sets of abstract values (such as closures or abstract numbers)
and a store.

\begin{lstlisting}
type VS = (Set[AbsValue], Store)
type Ans = R[Set[VS]]
\end{lstlisting}

Note that the type @Ans@ uses our stage polymorphic type @R@, meaning that under staging the type @Ans@
represents a next stage value. Then, our generic interpreter is also changed when handling function 
applications.

\begin{lstlisting}
case App(e1, e2) =>
  val e1ans = ev(e1, $\rho$, $\sigma$)
  val e1vs  = choices(e1ans)
  val e2ans = ev(e2, $\rho$, e1vs.$\sigma$)
  val e2vs  = choices(e2ans)
  apply_closure(ev)(e1vs.v, e2vs.v, e2vs.$\sigma$)
\end{lstlisting}

The idea is to handle the application is to explore all possible closures from @e1@ and meanwhile
use the latest store.
What @choices@ does is similar to McCarthy's @amb@ operator ~\cite{MCCARTHY196333}: it non-deterministically return an
element of type @VS@ from its argument, e.g., @e1ans@, to its right-hand side receiver, i.e., @e1vs@.
The function @choices@ internally uses delimited control operator @shift@ to capture the continuation,
which is the code block after its call site. 
This allows us to perform nesting depth-first evaluation for function application but still writing 
the program in direct-style \cite{Wei:2018:RAA:3243631.3236800}. 
Again, we can stage this part as we did for the naive abstract interpreter.

\subsubsection{Context-Sensitivity}

We add k-CFA-like context-sensitivity to the analysis by introducing an abstract time stamp, 
whose concrete instantiation is a finite list of expression that track $k$ recent calling context.
The definition of abstract addresses is changed to a tuple of identifiers and the time it get allocated,
meaning that this address points to some values under such calling context.
If $k$ is 0, we obtain a monovariant analysis as demonstrated before; if $k > 0$, a higher precision
analysis is obtained.

\begin{lstlisting}
type Time = List[Expr]
\end{lstlisting}

Every time when we call the @eval@ function, we refresh the time stamp by calling function @tick@,
which returns a new time stamp. Here we adopt a $k$-CFA-like allocation strategy, therefore
the @tick@ function can be implemented as appending the current expression being evaluated to the exiting
calling context, and then taking the first $k$ elements from the list.

\begin{lstlisting}
def tick(e: Expr, $\tau$: R[Time]): R[Time] = (e :: $\tau$).take(k)
def eval(ev: EvalFun)
  (e: Expr, $\rho$: R[Env], $\sigma$: R[Store], $\tau$: R[Time]): Ans = {
    val $\tau$_* = tick(e, $\tau$)
    e match {
      case Lit(i) => ...
      ...
    }
  }
\end{lstlisting}

Accordingly, the type of return value is accompanied with the time stamp. For recursive call of @ev@,
which would also return the latest time stamp, and that time stamp will be used for the afterward evaluation.

\begin{lstlisting}
type VST = (Value, Store, Time)
type Ans = R[Set[VST]]
\end{lstlisting}

Using other allocation strategies to achieve different sensitivities is possible \cite{DBLP:conf/icfp/Gilray0M16} 
and will be staged under our framework.

%\subsubsection{Mixed Sensitivity}

%using different $k$ for $k$-CFA

@@

Now we consider in a first-order imperative language, we may care more about the data-flow because the control-flow
is relatively easy to obtain. In this section, we show the staging of other abstract domains, particularly an 
interval domain for numbers.
It has been shown that specializing abstract domains with respect to the structure of analyzed program significantly 
improves the performance: a recent example is online decomposition of polyhedra \cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143}.
In this section, we first show how to support imperative language features to the abstract interpreter, 
and present a similar idea for interval domain but show the specialization is feasible by systematically staging.

@@

To evaluates an assignment, we first evaluate its right-hand side, and then put the value to the slot where
the address of @v@ points to.
For simplicity, we elect to make the value of the assignment be the value of its right-hand side.

\begin{lstlisting}
case Assign(x, e) =>
  val (v, $\sigma$_*) = ev(e, $\rho$, $\sigma$)
  (v, put($\sigma$_*, get($\rho$, x), v))
\end{lstlisting}

without introducing new language value members.

\begin{lstlisting}
case While(t, e) =>
  val (tv, t$\sigma$) = ev(t, $\rho$, $\sigma$)
  if (notZero(tv)) 
\end{lstlisting}

\subsubsection{Staged Interval}

\begin{lstlisting}
case class Interval(lb: Rep[Double], ub: Rep[Double]) {
  def +(that: Interval): Interval = that match {
    case Interval(lb_, ub_) => Interval(lb + lb_, ub + ub_)
  }
  def -(that: Interval): Interval = that match {
    case Interval(lb_, ub_) => Interval(lb - lb_, ub_ - ub) 
  }
  ...
}
\end{lstlisting}

### 2018-11-16 paper/70_related.tex
paper

\textbf{Abstract Compilation}. The idea in this paper is closely inspired by 
abstract compilation \cite{Boucher:1996:ACN:647473.727587}.
\citeauthor{Boucher:1996:ACN:647473.727587} presented abstract compilation
techniques as an efficient implementation of the monovariant flow analysis 
(\textit{0}-CFA) for programs written in continuation-passing style. 
The key idea is to remove the interpretation overhead on traversing the syntax 
tree by partial evaluation. Specifically, they proposed two similar kinds of abstract compilation techniques.
The first one is to generate specialized analysis as a textual program, then 
can be loaded and executed by \texttt{eval} or other similar mechanisms. 
The second one is using closure, i.e., a function that remembers its environment, 
as a representation of specialized analysis. 
So compiling the analysis generates higer-order functions on-the-fly with respect to the analyzed program,
then the generated closure can be applied immediately in the higher-order host language.

\citeauthor{Johnson:2013:OAA:2500365.2500604} adapt the idea closure generation 
for optimizing small-step abstract interpreter in state-transition style \cite{Johnson:2013:OAA:2500365.2500604}.
The analyzed program is firstly compiled to a intermediate representation called 
"abstract bytecode", which are actually higher-order functions, and then be executed 
on a abstract abstract machine for that IR.

%\citeauthor{DBLP:conf/flops/AmatoS01} apply abstract compilation for non par-sharing and freeness analysis in logic programs \cite{DBLP:conf/flops/AmatoS01}.

\citeauthor{damian1999partial}'s work on partial evaluation for program analysis \cite{damian1999partial}.

\citeauthor{amtoft1999partial}'s work on partial evaluation for constraint-based 
control flow analysis \cite{amtoft1999partial}.

\citeauthor{DBLP:conf/cgo/HardekopfL11} apply staging to flow-sensitive pointer analysis \cite{DBLP:conf/cgo/HardekopfL11}. 
They firstly stage the program code to obtain a sparse representation, then conduct
the flow-sensitive analysis.

\textbf{Abstract Interpreters}.

Abstract interpretation was proposed as a semantic-based approach to construct sound static analysis by approximation \cite{DBLP:conf/popl/CousotC77}.
As semantic artifacts, the Abstracting Abstract Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
approach shows that abstract interpreters can be derived systematically from concrete semantic artifacts.
The big-step abstract interpreters we presented in this paper are also inspired by the AAM framework,
which are later adopted for big-step abstract interpreters \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800}.

Polyhedra, online decomposition

\citet{DBLP:conf/cc/CousotC02} also proposed an abstract interpretation framework for modular analysis.
\citet{DBLP:conf/popl/CalcagnoDOY09} developed compositional shape analysis by using bi-abduction.

@@

Multi-stage programming. MetaML, MetaOCaml, Mint, LMS

Other notable implementations of MSP exist in ML family, for examples, MetaML \cite{DBLP:conf/pepm/TahaS97} 
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
Compared with the LMS approach in Scala, MetaML/MetaOCaml use term-level annotations such as brackets, 
escape, and run. Notwithstanding, we use LMS in this paper, the idea of staging an abstract interpreter
still applies with other MSP implementations.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization}
The idea in this paper is closely inspired by abstract compilation \cite{Boucher:1996:ACN:647473.727587}.
\citeauthor{Boucher:1996:ACN:647473.727587} presented abstract compilation
techniques as an efficient implementation of the monovariant flow analysis 
(\textit{0}-CFA) for programs written in continuation-passing style. 
The key idea is to remove the interpretation overhead on traversing the syntax 
tree by partial evaluation. Specifically, they proposed two similar kinds of abstract compilation techniques.
The first one is to generate specialized analysis as a textual program, which then
can be loaded and executed by \texttt{eval} or other similar mechanisms. 
The second one is to use closures, i.e., functions that remember their environments, 
as a representation of specialized analysis. As we show in the case study,
compiling the analysis generates higer-order functions on-the-fly with respect to 
the analyzed program, then the generated closure can be applied immediately in the 
higher-order host language.

\citeauthor{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure generation 
to optimize small-step abstract interpreter in state-transition style \cite{Johnson:2013:OAA:2500365.2500604}.
The analyzed program is firstly compiled to a intermediate representation called 
"abstract bytecode", which are actually higher-order functions, and then be executed 
on a abstract abstract machine for that IR.

\citet{damian1999partial} provides a formal treatment to abstract compilation and Shiver's CFA, 
as well as proofs to establish correctness and certified specialized analyzers \cite{damian1999partial}.
\citeauthor{amtoft1999partial} applyies partial evaluation for constraint-based 
control flow analysis \cite{amtoft1999partial}.
Split the analysis to be multiple stages is also studied other than control-flow analysis,
though the formulation may very different.
For example, \citeauthor{DBLP:conf/cgo/HardekopfL11} apply staging to flow-sensitive pointer 
analysis \cite{DBLP:conf/cgo/HardekopfL11}. The first stage is to analyze the program code to 
obtain a sparse representation, and then the second stage conducts the flow-sensitive analysis.

\citeauthor{DBLP:conf/popl/SinghPV17} optimizes polyhedra abstract domain through online decomposition \cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143}.
The idea is to decompose a large polyhedra into several smaller one; one abstract transformer
usually only depends a limit number of variables, then we just need to compute the resulting polyhera
on those smaller polyhedra that contains these related variables.
The decomposition can be considered as a specialization of abstract domain with respect to the program
static structure.

\paragraph{Abstract Interpreters}
Abstract interpretation was proposed as a semantic-based approach to construct sound static analysis by approximation \cite{DBLP:conf/popl/CousotC77}.
As semantic artifacts, the Abstracting Abstract Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
approach shows that abstract interpreters can be derived systematically from concrete semantic artifacts.
The big-step abstract interpreters we presented in this paper are also inspired by the AAM framework,
which are later adopted for big-step abstract interpreters \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800}.
\citet{DBLP:conf/cc/CousotC02} also proposed an abstract interpretation framework for modular analysis.
\citet{DBLP:conf/popl/CalcagnoDOY09} developed compositional shape analysis by using bi-abduction.
The big-step abstract interpreter we presented in this paper is inspired by \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800}.

@@

The Lightweight Modular Staging framework in Scala
Other notable implementations of MSP exist in ML family, for examples, MetaML \cite{DBLP:conf/pepm/TahaS97} 
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
Compared with the LMS approach in Scala, MetaML/MetaOCaml use term-level annotations such as brackets, 
escape, and run. Notwithstanding, we use LMS in this paper, the idea of staging an abstract interpreter
still applies with other MSP implementations.
